The forestecology R package for fitting and assessing neighborhood models of the effect of interspecific competition on the growth of trees

### 2021-03-10 paper/forestecology.Rmd
3060 -> 3000 🏁🏁🏁

We address this shortcoming with the `forestecology` R package providing methods and data for forest ecology model fitting and assessment available on CRAN (https://cran.r-project.org/web/packages/forestecology/index.html) and on GitHub (https://github.com/rudeboybert/forestecology). The package is written to model stem radial growth between two censuses based on neighborhood competition largely following the methods in @allen_permutation_2020.

@@

where $\beta_{0,j}$ is the diameter-independent growth rate for group $j$; $\text{dbh}_{ij}$ is the DBH of the focal tree at the earlier census for group $j$ and $\beta_{\text{dbh},j}$ quantifies its corresponding growth rate change;  $\text{BA}_{ijk}$ is the sum of the basal area of all trees of competitor species group $k$ and $\lambda_{jk}$ quantifies the corresponding change in growth for individuals of group $j$ from these competitors; and $\epsilon_{ij}$ is a random error term distributed $\text{Normal}(0, \sigma^2)$. They estimate all parameters via Bayesian linear regression while exploiting Normal/Inverse Gamma conjugacy to derive closed-form solutions to all posterior distributions^[See S1 Appendix of @allen_permutation_2020, available at https://doi.org/10.1371/journal.pone.0229930.s004]. These closed-form solutions are in contrast to approximations of all posteriors via computationally expensive Markov Chain Monte Carlo algorithms.

To evaluate whether competitor species identity matters, @allen_permutation_2020 run a permutation test where a null hypothesis of no species grouping-specific effects of competition is assumed, and thus the species identity of all competitors can be permuted:

@@

Furthermore, to account for the spatial autocorrelation of forestry data in their estimates of out-of-sample model error, @allen_permutation_2020 use spatial cross-validation. Estimates of model error that do not account for this spatial dependency tend to underestimate the true model error [@roberts_cross-validation_2017].

The package is designed with "tidy" design principles in mind [@wickham_welcome_2019]. Much like all `tidyverse` packages, `forestecology` has verb-named functions that can be modularly composed using the pipe `%>%` operator to sequentially complete all the necessary analysis steps [@bache_pipe_2020]. Furthermore, the inputs and outputs of most of our functions use the same "simple features for R" data structures from the `sf` package, a package for standardized and `tidyverse`-friendly wrangling and visualizing of spatial data [@pebesma_simple_2018]

@@

The package has the following ecological goals: to evaluate i) the effect of competitor species identity using permutation tests and ii) model performance using spatial cross-validation. We outline the four step basic analysis sequence:

@@

We start by loading all necessary packages.

@@

The first step is to compute the growth of trees using data from two censuses. `compute_growth()` computes the average annual growth based on census data that roughly follows ForestGEO standards. Despite such standards, minor variations will still exist between sites, thereby necessitating some data wrangling. For example, the SCBI site records all DBH values in millimeters [@bourg_initial_2013], whereas the Michigan Big Woods site used in @allen_permutation_2020 records them in centimeters [@allen_michigan_2020].

We first load both 2008 and 2014 SCBI census `.csv` files as they existed on GitHub on 2020/11/20 and perform minor data wrangling to both data sets [@gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020]. We then only consider a 9 ha subsection of the 25.6 ha of the SCBI site in order to speed up computation for this example: `gx` from 0--300 instead of 0--400 and `gy` from 300--600 instead of 0--640.

@@

These two data frames are then used as inputs to `compute_growth()`, along with the `id` argument specifying the variable that uniquely identifies each tree-stem. Note that we also discard all resprouts with `code == R` in the later census, since we are only interested in the growth of surviving, and not resprouted, stems.

@@

```{r}
# Manually define spatial blocks to act as folds
n_fold <- 4
fold1 <- rbind(c(0, 300), c(150, 300), c(150, 450), c(0, 450))
fold2 <- rbind(c(150, 300), c(300, 300), c(300, 450), c(150, 450))
fold3 <- rbind(c(0, 450), c(150, 450), c(150, 600), c(0, 600))
fold4 <- rbind(c(150, 450), c(300, 450), c(300, 600), c(150, 600))

@@

Lastly, we then fit the competition Bayesian linear regression model for tree growth outlined in Equation \ref{eq:model} using `comp_bayes_lm()`. This function has an option to specify prior distributions of all parameters of interest, chosen here to be the defaults detailed in `?comp_bayes_lm`.

@@

The resulting`comp_bayes_lm_scbi` is an object of S3 class type `comp_bayes_lm` containing the posterior values of all parameters. Furthermore, this class of object includes generics for three methods. First, the generic for `print()` displays the names of all prior and posterior parameters and the model formula:

@@

Next, the generic for `predict()` takes the posterior parameter values in `comp_bayes_lm_scbi` and a `newdata` data frame, and outputs a vector `growth_hat` of predicted values $\widehat{y}$ of the DBH computed from the posterior predictive distribution.

@@

We can now compare the observed and predicted growths to compute the root mean squared error (RMSE) of our model fit.

@@

Lastly, the generic for `ggplot2::autoplot()` allows us to visualize the posterior distribution of all parameters, as seen in Figure \ref{fig:scbi-posterior-viz}. Setting `type` to  `"intercepts"` and `"dbh_slopes"` returns species-specific posterior distributions for $\beta_{0,j}$ and $\beta_{dbh,j}$ respectively, while setting `type = "competition"` returns competition coefficients $\lambda_{j,k}$.

@@

For many users the visualizations relating to $\lambda_{j,k}$ will be of particular interest as they provide insight into species competitive interactions, where negative values indicate a competitor species which slows the growth of a focal species. Here, for example, we see that tulip poplars (litu) have a strong negative effect on the growth of conspecifics but relatively lesser effect on pignut hickory (cagl) and red oak (quru) neighbors.

Currently the `forestecology` package can only fit the competition Bayesian linear regression model outlined in Equation \ref{eq:model}. However, it can be extended to any model as long as it is implemented in a function similar to `comp_bayes_lm()`.

@@

To evaluate the effect of competitor species identity, we use the above four steps along with the permutation test in Equation \ref{eq:permutation-hypothesis-test}. Under a null hypothesis where competitor species identity does not matter, we can permute the competitor species identities within each focal tree, compute the RMSE test statistic, repeat this process several times to construct a null distribution, and finally compare it to the observed RMSE to assess significance. Going back to our example in Section \ref{focal-vs-comp} of focal tree with `focal_ID` 4 and its 20 competitors, the permutation test only randomly resamples the `comp_sp` variable without replacement, leaving all other variables intact. This resampling is nested within each focal tree in order to preserve neighborhood structure. We perform this permutation test, we once again use `comp_bayes_lm()` but by setting `run_shuffle = TRUE`.

@@

To evaluate model performance, we use spatial cross-validation. The example model fit in Section \ref{model-fit-predict} uses the same data to both fit the model and assess its performance. Given the inherent spatial-autocorrelation of our data, this can potentially lead to overfit models [@roberts_cross-validation_2017]. To mitigate the risk of overfitting, we make use of the spatial cross-validation blocking scheme encoded in the `foldID` variable from Section \ref{spatial-information} and visualized in Figure \ref{fig:scbi-spatial-information}.

At each iteration of our cross-validation, one fold will act as the test set and the remaining three acting as the training set. We fit the model to all focal trees in the training set, apply the model to all focal trees in the test set, compute predicted values, and compute the RMSE. Furthermore, to maintain spatial independence between the test and training set, a "fold buffer" that extend `r comp_dist`m outwards from the boundary of the test set is computed; all trees falling within this fold buffer are excluded from the training set (see Figure \ref{fig:scbi-spatial-cross-validation-schematic}).

@@

This is repeated for each of the four folds acting as the test set, then the four resulting RMSE's are averaged to provide a single estimate of model error. This algorithm is implemented in `run_cv()`, which acts as a wrapper function to both `comp_bayes_lm()` that fits the model and `predict()` that returns predicted values.

@@

The resulting RMSE of `r model_rmse_cv` computed using cross-validation is `r ifelse(model_rmse_cv > model_rmse, "larger", "smaller")` than the earlier RMSE of `r model_rmse`, suggesting that models that do not take spatial autocorrelation account generate model error estimates that are overly optimistic. In our case, RMSE values that are too low.

@@

`run_cv()` also accepts the `run_shuffle` argument in order to permute competitor species identity as described in Section \ref{permutation-test}. Figure \ref{fig:scbi-simulation} compares model performance for 49 permutations of competitor species and RMSE calculations, both with and without cross-validation. Without cross-validation, competitor species identity did matter as the observed RMSE was significantly lower than the permutation null distribution of RMSE. However, once we incorporate spatial cross-validation, this improvement disappears. These results suggest that in this 9 ha subplot of the SCBI plot competitive interactions do not depend on the identity of the competitor, which is the opposite of what has been observed in other locations [@allen_permutation_2020 @uriarte_spatially_2004]. Furthermore, the larger cross-validated RMSE's are indicative of the importance of being vigilant against potential model overfitting.

@@

The authors thank Sophie Li for their feedback on the package interface. The authors declare no conflicts of interest.

@@

We intend to archive all data and source code for this manuscript on GitHub at https://github.com/rudeboybert/forestecology. This repository will be archived on Zenodo upon acceptance. The example Smithsonian Conservation Biology Institute census data are available on GitHub at https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_main_census/data/census-csv-files and are archived on Zenodo at https://doi.org/10.5281/zenodo.2649301 [@gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020].

>>>>

We address this shortcoming with the `forestecology` R package providing methods and data for forest ecology model fitting and assessment, available on CRAN (https://cran.r-project.org/web/packages/forestecology/index.html) and on GitHub (https://github.com/rudeboybert/forestecology). The package is written to model stem radial growth between two censuses based on neighborhood competition, largely following the methods in @allen_permutation_2020.

@@

where $\beta_{0,j}$ is the diameter-independent growth rate for group $j$; $\text{dbh}_{ij}$ is the DBH of the focal tree at the earlier census for group $j$ and $\beta_{\text{dbh},j}$ quantifies its corresponding growth rate change;  $\text{BA}_{ijk}$ is the sum of the basal area of all trees of competitor species group $k$ and $\lambda_{jk}$ quantifies the corresponding change in growth for individuals of group $j$ from these competitors; and $\epsilon_{ij}$ is a random error term distributed $\text{Normal}(0, \sigma^2)$. They estimate all parameters via Bayesian linear regression, while exploiting Normal/Inverse Gamma conjugacy to derive closed-form solutions to all posterior distributions^[See S1 Appendix of @allen_permutation_2020, available at https://doi.org/10.1371/journal.pone.0229930.s004]. These closed-form solutions are not as computationally expensive as approximations from Markov Chain Monte Carlo algorithms.

To evaluate whether competitor species identity matters, @allen_permutation_2020 run a permutation test where a null hypothesis of no species grouping-specific effects of competition is assumed, thus the species identity of all competitors can be permuted:

@@

Furthermore, to account for the spatial autocorrelation in their estimates of out-of-sample model error, @allen_permutation_2020 use spatial cross-validation. Estimates of model error that do not account for this dependence tend to underestimate the true model error [@roberts_cross-validation_2017].

The package is designed with "tidy" design principles in mind [@wickham_welcome_2019]. Much like all `tidyverse` packages, `forestecology` has verb-named functions that can be modularly composed using the pipe `%>%` operator to sequentially complete all necessary analysis steps [@bache_pipe_2020]. Furthermore, the inputs and outputs of most functions use the same "simple features for R" data structures from the `sf` package, a package for standardized and `tidyverse`-friendly wrangling and visualizing of spatial data [@pebesma_simple_2018].

@@

The package has the following goals: to evaluate i) the effect of competitor species identity using permutation tests and ii) model performance using spatial cross-validation. We outline the four-step basic analysis sequence:

@@

We start by loading all packages.

@@

We first compute the growth of trees using data from two censuses. `compute_growth()` computes the average annual growth based on census data that roughly follows ForestGEO standards. Despite such standards, minor variations will still exist between sites, thereby necessitating some data wrangling. For example, the SCBI site records all DBH values in millimeters [@bourg_initial_2013], whereas the Michigan Big Woods site used in @allen_permutation_2020 records them in centimeters [@allen_michigan_2020].

We load both 2008 and 2014 SCBI census `.csv` files as they existed on GitHub on 2020/11/20 and perform minor data wrangling [@gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020]. We then only consider a 9 ha subsection of the 25.6 ha of the site to speed up computation for this example: `gx` from 0--300 instead of 0--400 and `gy` from 300--600 instead of 0--640.

@@

These two data frames are then used as inputs to `compute_growth()`, along with `id` specifying the variable that uniquely identifies each tree-stem. We also discard all resprouts with `code == R` in the later census, since we are only interested in the growth of surviving, and not resprouted, stems.

@@

```{r}
# Manually define spatial blocks to act as folds
n_fold <- 4
fold1 <- cbind(c(0, 150, 150, 0), c(300, 300, 450, 450))
fold2 <- cbind(c(150, 300, 300, 150), c(300, 300, 450, 450))
fold3 <- cbind(c(0, 150, 150, 0), c(450, 450, 600, 600))
fold4 <- cbind(c(150, 300, 300, 150), c(450, 450, 600, 600))

@@

Lastly, we fit the competition Bayesian linear regression model for tree growth outlined in Equation \ref{eq:model} using `comp_bayes_lm()`. This function has an option to specify prior distributions of all parameters, chosen here to be the defaults detailed in `?comp_bayes_lm`.

@@

The resulting`comp_bayes_lm_scbi` is an object of S3 class type `comp_bayes_lm` containing the posterior values of all parameters. Furthermore, this class includes generics for three methods. First, the generic for `print()` displays the names of all prior and posterior parameters and the model formula:

@@

Next, the generic for `predict()` takes the posterior parameter values in `comp_bayes_lm_scbi` and a `newdata` data frame, and outputs a vector `growth_hat` of predicted DBH values $\widehat{y_{ij}}$ computed from the posterior predictive distribution.

@@

We can now compare the observed and predicted growths to compute the root mean squared error (RMSE) of our model:

@@

Lastly, the generic for `ggplot2::autoplot()` allows us to visualize all posterior distributions, as seen in Figure \ref{fig:scbi-posterior-viz}. Setting `type` to `"intercepts"` and `"dbh_slopes"` returns species-specific posterior distributions for $\beta_{0,j}$ and $\beta_{dbh,j}$ respectively, while setting `type = "competition"` returns competition coefficients $\lambda_{j,k}$.

@@

For many users the visualizations of $\lambda_{j,k}$ will be of particular interest as they provide insight into species competitive interactions, where negative values indicate a competitor species which slows the growth of a focal species. Here, for example, we see that tulip poplars (litu) have a strong negative effect on the growth of conspecifics but relatively lesser effect on pignut hickory (cagl) and red oak (quru) neighbors.

Currently the `forestecology` package can only fit the competition Bayesian linear regression model in Equation \ref{eq:model}. However, it can be extended to any model as long as it is implemented in a function similar to `comp_bayes_lm()`.

@@

To evaluate the effect of competitor species identity, we use the above four steps along with the permutation test in Equation \ref{eq:permutation-hypothesis-test}. Under a null hypothesis where competitor species identity does not matter, we can permute the competitor species identities within each focal tree, compute the RMSE test statistic, repeat this process several times to construct a null distribution, and compare it to the observed RMSE to assess significance. Going back to our example in Section \ref{focal-vs-comp} of focal tree with `focal_ID` 4 and its 20 competitors, the permutation test only randomly resamples the `comp_sp` variable without replacement, leaving all other variables intact. This resampling is nested within each focal tree in order to preserve neighborhood structure. We perform this permutation test once again using `comp_bayes_lm()` but by setting `run_shuffle = TRUE`.

@@

To evaluate model performance, we use spatial cross-validation. The model fit in Section \ref{model-fit-predict} uses the same data to both fit and assess model performance. Given the spatial-autocorrelation of our data, this can potentially lead to overfit models [@roberts_cross-validation_2017]. To mitigate this risk, we use the spatial cross-validation blocking scheme encoded in the `foldID` variable from Section \ref{spatial-information} and visualized in Figure \ref{fig:scbi-spatial-information}.

At each iteration of the cross-validation, one fold acts as the test set and the remaining three act as the training set. We fit the model to all focal trees in the training set, apply the model to all focal trees in the test set, compute predicted values, and compute the RMSE. Furthermore, to maintain spatial independence between the test and training sets, a "fold buffer" that extends `r comp_dist`m outwards from the boundary of the test set is considered; all trees within this "fold buffer" are excluded from the training set (see Figure \ref{fig:scbi-spatial-cross-validation-schematic}).

@@

This process is repeated for each of the four folds acting as the test set, then the four RMSE's are averaged to provide a single estimate of model error. This algorithm is implemented in `run_cv()`, which acts as a wrapper function to both `comp_bayes_lm()` that fits the model and `predict()` that returns predicted values.

@@

The resulting RMSE of `r model_rmse_cv` computed using cross-validation is `r ifelse(model_rmse_cv > model_rmse, "larger", "smaller")` than the earlier RMSE of `r model_rmse`, suggesting that models that do not account for spatial autocorrelation generate model error estimates that are overly optimistic, i.e. RMSE values that are too low.

@@

`run_cv()` also accepts the `run_shuffle` argument in order to permute competitor species identity as described in Section \ref{permutation-test}. Figure \ref{fig:scbi-simulation} compares model performance for 49 permutations of competitor species and RMSE calculations, both with and without cross-validation. Without cross-validation, competitor species identity does matter as the observed RMSE was significantly lower than the permutation null distribution of RMSE. However, once we incorporate spatial cross-validation, this improvement disappears. These results suggest that in this 9 ha subplot of the SCBI plot, competitive interactions do not depend on the identity of the competitor, which is the opposite of what has been observed in other locations [@allen_permutation_2020 @uriarte_spatially_2004]. Furthermore, the larger cross-validated RMSE's are indicative of the importance of being vigilant against potential model overfitting.

@@

We thank Sophie Li for their feedback on the package interface. The authors declare no conflicts of interest.

@@

We intend to archive all data and source code for this manuscript on GitHub at https://github.com/rudeboybert/forestecology and on Zenodo upon acceptance. The example Smithsonian Conservation Biology Institute census data are available on GitHub at https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_main_census/data/census-csv-files and are archived on Zenodo at https://doi.org/10.5281/zenodo.2649301 [@gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020].
